Multi-stage programming with functors and monads: eliminating abstraction overhead from generic code

### 2007-07-11 SCP2008/paper/metamonads.tex
finish subsection on 'organizing aspects'

@@

There is an interesting relation with aspect-oriented code
\cite{kiczales97aspectoriented}: in AspectJ, 
aspects are (comparatively) lightly typed, and are post-facto extensions of an
existing piece of code.  Here aspects are weaved together ``from scratch'' to
make up a piece of code/functionality.  One can understand previous work to be
more akin to dynamically typed aspect weaving, while we have started
investigating statically typed one.

>>>>

Further investigation of the aspects reveals that these fall into
various categories.  This categorization has already been useful to us,
as it revealed that parts of our implementation was not as well
separated into distinct concerns as we had originally thought.  In
particular, it revealed that our implementation of ``containers'' 
mixed representation (i.e. data-structure) issues with abstraction
issues (i.e. matrices and their natural representation-independent
operations).  We hope to rectify this in the future.

\begin{enumerate}
	\item \textbf{Abstract domain}.  In other words, these are
		firmly aspects which are domain-specific, and encode
		requirements level entities.  In particular, items like
		Domain, length measure, Matrix, permutation, full division
		and structured matrices all belong here.
	\item \textbf{Concrete representation}.  In other words, the
		design-time issue of choosing a data-structure to represent
		various abstract domains.  Items like whether the results 
		should be packed, (part of) the representation of containers,
		sparsity, and representation of permutations.
	\item \textbf{Interface}.  In other words, what will be the
		interface of each generated function.  This mostly deals
		with the input and output choices, but also includes whether
		the output should be in-place, and whether errors or warnings
		should be issued in certain situations.  Logging is also
		an interface issue.
	\item \textbf{Algorithmic Strategy}.  This encodes the various
		choices in the algorithms themselves.  For example, whether
		to use fraction-free updates, which pivot strategy to use,
		whether the matrix is augmented.
	\item \textbf{Information tracking}.  What information should be
		kept and/or computed while the algorithm proceeds, such
		as whether to track determinant, rank, or pivot.
	\item \textbf{Staging}.  This is the abstract interface to the
		underlying programming language.  It (currently) encodes
		whether to generate code that will be executed in a future
		stage, or whether the code is direct and should be executed
		now.  It should be possible to encode other strategies too.
\end{enumerate}

While the first two families of aspects are the most obvious, it is quite
difficult to separate them out cleanly.  This seems to stem from the fact
that in a non-staged setting, separating out these aspects leads to such
fantastically inefficient code, especially in the setting of scientific
computation, that our experience works \emph{against us}.  While it would
be too much to speak of a paradigm shift, it is necessary to firmly think
in terms of abstract ideal design while avoiding all thoughts of traditional
efficiency.

Interestingly, unlike traditional approaches \cite{journals/cacm/parnas72a}, we
allow the interfaces to our (generated) routines to vary.  Compared to dynamic
approaches in Object-Oriented language (late binding and dynamic dispatch) or
functional languages (Haskell's dictionary-based \emph{type classes}), which
offer some flexibility on interfaces, our interfaces are actually fixed at
run-time, thus do incur any run-time inefficiency whatsoever.

While we are by no means the first to speak of algorithmic strategy, to
our knowledge our approach is one of the cleanest, in that the issue of
strategy is completely separated into an independent entity (module), and
yet we suffer no run-time inefficiency because of it.

The information tracking aspects are in some ways to most intruiging ones:
they are purely internal to the algorithms, and they are the ones which
are most dependent on other choices.  The \emph{implementation} of these
aspects is completely orthogonal to the other aspects, however their
\emph{compositionality properties} are deeply enmeshed with that of other
aspects.  While writing the code for the aspects was straightforward enough,
this is where quite a bit of infrastructure was needed to encode
various dependencies, so that it would be impossible to generate an
instance of the LU Decomposition algorithm with information
tracking aspects which were incompatible with the requirements of other
aspects.

Lastly, to ensure that we were indeed firmly in a \emph{cogen} setting,
we abstracted out the underlying programming language completely.  This 
allowed us to both generate (efficient) code and to write a directly
runnable (highly inefficient) version of the algorithm from the same
generator.  This brought out the issue that, once this was done, we had
essentially abstracted out all of the syntactic sugar of the underlying
programming language, and all we were left with was function application
and monadic composition.  This means the code for our generator looks mostly
like Scheme with added syntax for monads!

@@

\jacques{Mention the relationship with the all-dynamic version of a 
similar design, namely Maple's?  The fact that all choices there are
dynamic, while all ours are completely static really shows that the
design space along the dynamic--static dimension is quite large and
versatile.}

There is an interesting relation with aspect-oriented code
\cite{kiczales97aspectoriented}: in AspectJ, 
aspects are (comparatively) lightly typed, and are post-facto extensions of an
existing piece of code.  More precisely, they are post-facto extensions of
potential program traces, specified in a particular language; these tend
to be created to follow the operational behaviour of existing code, but
is not restricted to such a setting.
In our work, aspects are weaved together ``from scratch'' to
make up a piece of code/functionality.  One can understand previous work to be
more akin to dynamically typed, and dynamically specified aspect weaving, while
we have started investigating statically typed and statically specified
aspect weaving.
